---
title: "PilottoAquaSYNC"
output: html_document
date: "2026-03-12"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
install.packages('arrow') # for reading .parquet files
library(arrow)
library(tidyverse)
install.packages('codyn')
library(codyn)
library(data.table)
```

```{r}
setwd("~/Documents/FEM_work/Pilotto/AquaSYNC_Pilotto")
```


```{r}
theme_set(theme_bw())
```

# Initial import - just to extract the coordinates
```{r}

benthos<-
read_parquet('BenthicMacroinvertebrates.parquet')
```

```{r}
benthos_coords<-
  benthos[,c(5,8,9)]


write.csv(benthos_coords, 'benthos_coords.csv')
```



#########################################
##### import the taxon cleaned macroinvertebrate data, for initial analyses
##############################

```{r}
benthos_family<-
read_parquet('BenthicMacroinvertebrates_Family_and_higher.parquet')
```



# there are issues with SampleID (some are just numbers, not sure it's ok ).
```{r}
table(benthos_family$SiteID)
table(benthos_family$SampleID)

```

# convert date to Julian
```{r}
benthos_family$Date_julian<-
  as.numeric(benthos_family$Date)
  
```

# some checking
```{r}
tmp<-
  benthos_family %>% 
  filter(Country=='Norway')

unique(tmp$SiteID)

```
# the synchrony works on Norway
```{r}
codyn::synchrony(tmp, time.var = 'Date_julian', species.var = 'Taxon.harm', abundance.var = 'Value', replicate.var = 'SiteID')
```

# get number of unique taxa and unique date for siteID
#also get richness per site/time combination
```{r}
#how many taxa per site
Site_taxa_counts<-
benthos_family %>% 
  group_by(SiteID) %>% 
  summarise(taxa_count= n_distinct(Taxon.harm)) %>% 
  arrange(taxa_count)

#how many observations per site (not necessarily years, but just samples)
Site_date_counts<-
benthos_family %>% 
  group_by(SiteID) %>% 
  summarise(date_count= n_distinct(Date_julian)) %>% 
  arrange(date_count)

#richness per site / time combination
Site_taxa_counts_Date<-
benthos_family %>% 
  group_by(SiteID, Date) %>% 
  summarise(taxa_rich= n_distinct(Taxon.harm)) 

Site_taxa_counts_Date$Country<-
  benthos_family$Country[match(Site_taxa_counts_Date$SiteID, benthos_family$SiteID)]


  
```

# For curisiosity check if trends in richness over time (for country)
```{r}
Site_taxa_counts_Date %>% 
  mutate(Date_julian=as.numeric(Date)) %>% 
  ggplot() +aes(Date_julian, taxa_rich, col=Country)+geom_smooth(method='lm',se=T)
```


```{r}
Site_taxa_counts_Date %>% 
  filter(SiteID %in% Site_date_counts[Site_date_counts$date_count>10,]$SiteID) %>% 
  mutate(Date_julian=as.numeric(Date)) %>% 
  ggplot() +aes(Date_julian, taxa_rich, group=SiteID, col=Country)+geom_smooth(method='lm',se=F, size=0.2)



Site_taxa_counts_Date %>% 
  filter(SiteID %in% Site_date_counts[Site_date_counts$date_count>10,]$SiteID) %>% 
  mutate(Date_julian=as.numeric(Date)) %>% 
  ggplot() +aes(Date_julian, taxa_rich, group=SiteID, col=Country)+geom_smooth(method='lm',se=F, size=0.2)+facet_wrap(~Country)

```




# some sites have only one time point
#example of sites with no temporal observation
```{r}
benthos_family %>% 
  filter(SiteID=="SE655380-155738|sublittoral")


benthos_family %>% 
  filter(SiteID=="SE629026-147562|sublittoral")


```

#subset based on only siteID with >1 time observation
#Also remove chordata
*could better subset sites >10 observation*
```{r}
benthos_family_subs<-
benthos_family %>% 
  filter(SiteID %in% Site_date_counts[Site_date_counts$date_count>1,]$SiteID) %>% 
  filter(phylum !='Chordata')

```


# first synchrony run shows issues with replicated values
*copy the error message below*
```{r}
codyn::synchrony(benthos_family_subs, time.var = 'Date_julian', species.var = 'Taxon.harm', abundance.var = 'Value', replicate.var = 'SiteID')
```

Error in check_single(df, time.var, species.var, replicate.var) : 
  Multiple records for one or more species found at:
 Taxon.harm          SiteID       
 "Ameletidae"        "BEA01_a"    
 "Baetidae"          "BEA01_a"    
 "Capniidae"         "BEA01_a"    
 "Chironomidae"      "BEA01_a"    
 "Chloroperlidae"    "BEA01_a"    
 "Enchytraeidae"     "BEA01_a"    
 "Ephemerellidae"    "BEA01_a"    
 "Glossosomatidae"   "BEA01_a"    
 "Heptageniidae"     "BEA01_a"    
 "Hydropsychidae"    "BEA01_a"    
 "Lepidostomatidae"  "BEA01_a"    
 "Naididae"          "BEA01_a"    
 "Nemouridae"        "BEA01_a"    
 "Pediciidae"        "BEA01_a"    
 "Perlodidae"        "BEA01_a"    
 "Planariidae"       "BEA01_a"    
 "Rhyacophilidae"    "BEA01_a"    
 "Taeniopterygidae"  "BEA01_a"    
 "Ameletidae"        "BEL01_c"    
 "Baetidae"          "BEL01_c"    
 "Capniidae"         "BEL01_c"    
 "Chironomidae"      "BEL01_c"    
 "Chloroperlidae"    "BEL01_c"    
 "Collembola"        "BEL01_c"    
 "Dixidae"           "BEL01_c"    
 "Elmidae"           "BEL01_c"



# check the problems with duplicated data in some Canada sites
*to idetify duplicates, I exlcude the 'sample_id' column, which differs for duplicated entries*
#e.g. the BEA01_a site

*some species have duplicated entries, with identical abundance values and dates 2, or 3 times*
```{r}
BEA01_a<-
benthos_family_subs %>% 
  filter(SiteID =="BEA01_a")

xx<-BEA01_a[duplicated(BEA01_a[,-10]) |duplicated(BEA01_a[,-10], fromLast = T)  ,]


benthos_family_subs %>% 
  filter(SiteID =="BX01_b")

```

# extract all duplicated data
```{r}
dupli<-benthos_family_subs[duplicated(benthos_family_subs[,-10]) |duplicated(benthos_family_subs[,-10], fromLast = T)  ,]
```

# remove all duplicates (but keep the first entry)
```{r}
benthos_family_subs[,-10] %>% 
  distinct()
 
```


# extract the Loreau and Gross synchrony (community synch), for the cleaned benthos df
*(no duplicates, no sites w/o time series)*
```{r}
benthos_fam_synchrony<-
codyn::synchrony(benthos_family_subs[,-10] %>% distinct(),
                 
                 time.var = 'Date_julian', species.var = 'Taxon.harm', abundance.var = 'Value', replicate.var = 'SiteID')

benthos_fam_synchrony$Gross<-
  codyn::synchrony(benthos_family_subs[,-10] %>% distinct(),
                 
                 time.var = 'Date_julian', species.var = 'Taxon.harm', abundance.var = 'Value', replicate.var = 'SiteID', metric = 'Gross')$synchrony

```

# add some covariates to sites synchrony
```{r}
benthos_fam_synchrony$Lon<-
  benthos_family_subs$Lon[match(benthos_fam_synchrony$SiteID, benthos_family_subs$SiteID)]

benthos_fam_synchrony$Lat<-
  benthos_family_subs$Lat[match(benthos_fam_synchrony$SiteID, benthos_family_subs$SiteID)]

benthos_fam_synchrony$n_time<-
  Site_date_counts$date_count[match(benthos_fam_synchrony$SiteID, Site_date_counts$SiteID)]

benthos_fam_synchrony$n_taxa<-
  Site_taxa_counts$taxa_count [match(benthos_fam_synchrony$SiteID, Site_taxa_counts$SiteID)]

benthos_fam_synchrony$Country<-
    benthos_family_subs$Country [match(benthos_fam_synchrony$SiteID, benthos_family_subs$SiteID)]

```


# Explore relationship between comm synchrony and n_taxa, n_time, Lat
```{r}
benthos_fam_synchrony %>% 
  ggplot()+aes(Lat, synchrony, col=Country)+geom_point()+geom_smooth(se=F)

benthos_fam_synchrony %>% 
  ggplot()+aes(Lat, synchrony, col=Country)+geom_point()+geom_smooth(se=F, method='lm')


benthos_fam_synchrony %>% 
  ggplot()+aes(n_taxa, synchrony, col=Country)+geom_point()+geom_smooth(se=F, method='lm')

benthos_fam_synchrony %>% 
  ggplot()+aes(n_taxa, Gross, col=Country)+geom_point()+geom_smooth(se=F, method='lm')


benthos_fam_synchrony %>% 
  ggplot()+aes(n_time, synchrony, col=Country)+geom_point()+geom_smooth(se=F, method='lm')

benthos_fam_synchrony %>% 
  ggplot()+aes(n_time, Gross, col=Country)+geom_point()+geom_smooth(se=F, method='lm')


benthos_fam_synchrony %>% 
  ggplot()+aes(Lat, n_taxa, col=Country)+geom_point()+geom_smooth(se=F, method='lm')

```


# Extract year vector
# create *list with 6y time window*
```{r}
yearID<-sort(unique(benthos_family_subs$Year))


time.windowlist=list()
ywindow=5
for(i in 1:(length(yearID)-ywindow)){
time.windowlist[[i]]=  (yearID[i:(i+5)])}

```

# how to subset the general data with specific year window
```{r}
benthos_family_subs %>% 
  filter(Year %in% time.windowlist[[1]])
```

# create time window ID vector (maybe for plotting)
```{r}

tw_ID<-rep(NA, 33)

for(tw in 1:length(time.windowlist)){
  tw_ID[tw]<-

  paste(time.windowlist[[tw]][1], time.windowlist[[tw]][6], sep="_")
}
```



# create a SiteID vector (overall)
```{r}
SiteID_vector<-unique(benthos_family_subs$SiteID)
```

# site ID vector for sites with >10 observations
```{r}
SiteID_10y<-unique(benthos_family_subs %>% 
                     filter(SiteID %in% Site_date_counts[Site_date_counts$date_count>10,]$SiteID) %>%
                     pull(SiteID)
)
```




# Prove Synchrony on given time window, filtering and selecting only sites with >4 years in given window
```{r}

codyn::synchrony(
benthos_family_subs[,-10] %>% 
                   distinct() %>% 
                  filter(SiteID %in% Site_date_counts[Site_date_counts$date_count>10,]$SiteID) %>% 
                   filter(Year %in% time.windowlist[[26]]) %>% 
                  group_by(SiteID) %>% 
                  mutate(nyears= n_distinct(Year)) %>% 
                  filter(nyears>4) %>% 
                  group_by(SiteID) %>% 
                  mutate(count_taxa=n_distinct(Taxon.harm)) %>% 
                  filter(count_taxa>6)
,

time.var = 'Date_julian', species.var = 'Taxon.harm', abundance.var = 'Value', replicate.var = 'SiteID')

```

# create list holding the moving window synch calculation
```{r}
synch.list<-list()
```

```{r}
for (tw in 1: length(time.windowlist)) {
  synch.list[[tw]]<-
    
    codyn::synchrony(
benthos_family_subs[,-10] %>% 
                   distinct() %>% # this exclude replicates (canada)
                  filter(SiteID %in% Site_date_counts[Site_date_counts$date_count>10,]$SiteID) %>% # sites with > 10 observations
                   filter(Year %in% time.windowlist[[tw]]) %>% # select given time window
                  group_by(SiteID) %>% 
                  mutate(nyears= n_distinct(Year)) %>% 
                  filter(nyears>4) %>% # at least 4 years in a given window (of 6)
                  group_by(SiteID) %>% 
                  mutate(count_taxa=n_distinct(Taxon.harm)) %>% 
                  filter(count_taxa>6) # sites with at least 6 taxa
,

time.var = 'Date_julian', species.var = 'Taxon.harm', abundance.var = 'Value', replicate.var = 'SiteID')
}
```


```{r}
synch.list[2]
```

# combine elements of list *using reduce (like cbind)*, keeping all siteID and adding NA when a given site has no value in a given timewindow. Not all sites are present in each element of the list. Using reduce we progressively cbind the data keeping all sites, and when a given site is absent (for a given time window) it will produce NAs
```{r}
synch_mw_test<- # call it as a test for now
synch.list %>% 
reduce(full_join, by = "SiteID")

colnames(synch_mw_test)[2:ncol(synch_mw_test)]<-tw_ID
```

# add Country to the synchrony time window
```{r}
synch_mw_test$Country<-
  benthos_family_subs$Country[match(synch_mw_test$SiteID, benthos_family_subs$SiteID)]

(synch_mw_test$Country)
```

# Transform the synch mw in long format (for plotting)
```{r}
synch_mw_test_long<-
synch_mw_test %>% 
  pivot_longer(cols = -c(SiteID, Country), names_to = 'Year_window', values_to = 'Com.Synchrony')
```

```{r}
synch_mw_test_long %>% 
  group_by(SiteID) %>% 
  mutate(na_count = sum(is.na(Com.Synchrony))) %>% #count the NA (for the sites with now synch in given window)
  filter(na_count<20) %>% # exclude sites with > n windows missing
  #mutate(Year_window_n=as.numeric(as.factor(Year_window))) %>% 
  separate(Year_window, into = c("start", "end"), sep = "_", convert = TRUE) %>% 
  mutate(Year_Midpoint = (start + end) / 2) %>% 
  ggplot()+aes(Year_Midpoint, Com.Synchrony, col=Country)+geom_smooth(se=F)
```

```{r}
synch_mw_test_long %>% 
  group_by(SiteID) %>% 
  mutate(na_count = sum(is.na(Com.Synchrony))) %>% #count the NA (for the sites with now synch in given window)
  filter(na_count<20) %>% # exclude sites with > n windows missing
  #mutate(Year_window_n=as.numeric(as.factor(Year_window))) %>% 
  separate(Year_window, into = c("start", "end"), sep = "_", convert = TRUE) %>% 
  mutate(Year_Midpoint = (start + end) / 2) %>% 
  ggplot()+aes(Year_Midpoint, Com.Synchrony, col=Country)+geom_smooth(aes(group=SiteID), se=F, size=0.1)+
  geom_smooth(aes(group=Country), se=F)
```

