---
title: "PilottoAquaSYNC"
output: html_document
date: "2026-03-12"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
install.packages('arrow') # for reading .parquet files
library(arrow)
library(tidyverse)
install.packages('codyn')
library(codyn)
library(data.table)

install.packages('geosphere') # for geog distance
library(geosphere)

install.packages('widyr') # for large sparce matrices / correlation
library(widyr)


library(iCAMP)
```

```{r}
setwd("~/Documents/FEM_work/Pilotto/AquaSYNC_Pilotto")
```


```{r}
theme_set(theme_bw())
```

# Initial import - just to extract the coordinates
```{r}

benthos<-
read_parquet('BenthicMacroinvertebrates.parquet')
```

```{r}
benthos_coords<-
  benthos[,c(5,8,9)]


write.csv(benthos_coords, 'benthos_coords.csv')
```



#########################################
##### import the taxon cleaned macroinvertebrate data, for initial analyses
##############################

```{r}
benthos_family<-
read_parquet('BenthicMacroinvertebrates_Family_and_higher.parquet')


benthos_clean<-
read_parquet('BenthicMacroinvertebrates_HarmonizedTaxonomy_SiteLevel.parquet')
```

#

# there are issues with SampleID (some are just numbers, not sure it's ok ).
```{r}
table(benthos_family$SiteID)
table(benthos_family$SampleID)

```

# convert date to Julian
```{r}
benthos_family$Date_julian<-
  as.numeric(benthos_family$Date)
  
benthos_clean$Date_julian<-
  as.numeric(benthos_clean$Date)

```

# some checking
```{r}
tmp<-
  benthos_clean %>% 
  filter(Country=='Norway')

unique(tmp$SiteID)


xx<-
tmp %>% 
  filter(SiteID=='2380')
```

# the synchrony works on Norway
```{r}
codyn::synchrony(tmp, time.var = 'Date_julian', species.var = 'Taxon.harm', abundance.var = 'Value', replicate.var = 'SiteID')
```

# get number of unique taxa and unique date for siteID
#also get richness per site/time combination
```{r}
#how many taxa per site
Site_taxa_counts<-
benthos_clean %>% 
  group_by(SiteID) %>% 
  summarise(taxa_count= n_distinct(Taxon.harm)) %>% 
  arrange(taxa_count)

#how many observations per site (not necessarily years, but just samples)
Site_date_counts<-
benthos_clean %>% 
  group_by(SiteID) %>% 
  summarise(date_count= n_distinct(Date_julian)) %>% 
  arrange(date_count)

#richness per site / time combination
Site_taxa_counts_Date<-
benthos_clean %>% 
  group_by(SiteID, Date) %>% 
  summarise(taxa_rich= n_distinct(Taxon.harm)) 

Site_taxa_counts_Date$Country<-
  benthos_clean$Country[match(Site_taxa_counts_Date$SiteID, benthos_clean$SiteID)]


  
```

# For curisiosity check if trends in richness over time (for country)
```{r}
Site_taxa_counts_Date %>% 
  mutate(Date_julian=as.numeric(Date)) %>% 
  ggplot() +aes(Date_julian, taxa_rich, col=Country)+geom_smooth(method='lm',se=T)
```


```{r}
Site_taxa_counts_Date %>% 
  filter(SiteID %in% Site_date_counts[Site_date_counts$date_count>10,]$SiteID) %>% 
  mutate(Date_julian=as.numeric(Date)) %>% 
  ggplot() +aes(Date_julian, taxa_rich, group=SiteID, col=Country)+geom_smooth(method='lm',se=F, size=0.2)



Site_taxa_counts_Date %>% 
  filter(SiteID %in% Site_date_counts[Site_date_counts$date_count>10,]$SiteID) %>% 
  mutate(Date_julian=as.numeric(Date)) %>% 
  ggplot() +aes(Date_julian, taxa_rich, group=SiteID, col=Country)+geom_smooth(method='lm',se=F, size=0.2)+facet_wrap(~Country)

```




# some sites have only one time point
#example of sites with no temporal observation
```{r}
benthos_family %>% 
  filter(SiteID=="SE655380-155738|sublittoral")


benthos_family %>% 
  filter(SiteID=="SE629026-147562|sublittoral")


```

#subset based on only siteID with >3 time observation
#Also remove chordata
*could better subset sites >10 observation*
```{r}
benthos_clean_subs<-
benthos_clean %>% 
  filter(SiteID %in% Site_date_counts[Site_date_counts$date_count>3,]$SiteID) %>% 
  filter(phylum !='Chordata')

```


# first synchrony run shows issues with replicated values
*copy the error message below*
```{r}
codyn::synchrony(benthos_clean_subs, time.var = 'Date_julian', species.var = 'Taxon.harm', abundance.var = 'Value', replicate.var = 'SiteID')
```






# extract all duplicated data
```{r}
dupli<-benthos_clean_subs[duplicated(benthos_clean_subs[,-10]) |duplicated(benthos_clean_subs[,-10], fromLast = T)  ,]
```

# *remove all duplicates (but keep the first entry)*
*This is left-over from previous run, now there are no duplicated rows*
```{r}
benthos_family_subs[,-10] %>% 
  distinct()
 
```


# extract the Loreau and Gross synchrony (community synch), for the cleaned benthos df
*(no duplicates, no sites w/o time series)*
```{r}
benthos_clean_synchrony<-
codyn::synchrony(benthos_clean_subs[,-10],
                 
                 time.var = 'Date_julian', species.var = 'Taxon.harm', abundance.var = 'Value', replicate.var = 'SiteID')

benthos_clean_synchrony$Gross<-
  codyn::synchrony(benthos_clean_subs[,-10] %>% distinct(),
                 
                 time.var = 'Date_julian', species.var = 'Taxon.harm', abundance.var = 'Value', replicate.var = 'SiteID', metric = 'Gross')$synchrony

```

# add some covariates to sites synchrony
```{r}
benthos_clean_synchrony$Lon<-
  benthos_clean_subs$Lon[match(benthos_clean_synchrony$SiteID, benthos_clean_subs$SiteID)]

benthos_clean_synchrony$Lat<-
  benthos_clean_subs$Lat[match(benthos_clean_synchrony$SiteID, benthos_clean_subs$SiteID)]

benthos_clean_synchrony$n_time<-
  Site_date_counts$date_count[match(benthos_clean_synchrony$SiteID, Site_date_counts$SiteID)]

benthos_clean_synchrony$n_taxa<-
  Site_taxa_counts$taxa_count [match(benthos_clean_synchrony$SiteID, Site_taxa_counts$SiteID)]

benthos_clean_synchrony$Country<-
    benthos_clean_subs$Country [match(benthos_clean_synchrony$SiteID, benthos_clean_subs$SiteID)]

benthos_clean_synchrony$LakeRiver<-
     benthos_clean_subs$Lake.river[match(benthos_clean_synchrony$SiteID, benthos_clean_subs$SiteID)]

```


# Explore relationship between comm synchrony and n_taxa, n_time, Lat
```{r}


benthos_clean_synchrony %>% 
  ggplot()+aes(Lat, synchrony, col=Country)+geom_point()+geom_smooth(se=F, method='lm')


benthos_clean_synchrony %>% 
  ggplot()+aes(n_taxa, synchrony, col=Country)+geom_point()+geom_smooth(se=F, method='lm')

benthos_clean_synchrony %>% 
  ggplot()+aes(n_taxa, Gross, col=Country)+geom_point()+geom_smooth(se=F, method='lm')


benthos_clean_synchrony %>% 
  ggplot()+aes(n_time, synchrony, col=Country)+geom_point()+geom_smooth(se=F, method='lm')

benthos_clean_synchrony %>% 
  ggplot()+aes(n_time, Gross, col=Country)+geom_point()+geom_smooth(se=F, method='lm')


benthos_clean_synchrony %>% 
  ggplot()+aes(Lat, n_taxa, col=Country)+geom_point()+geom_smooth(se=F, method='lm')


benthos_clean_synchrony %>% 
  ggplot()+aes(LakeRiver, synchrony)+geom_violin()+geom_boxplot()

```


# Extract year vector
# create *list with 6y time window*
```{r}
yearID<-sort(unique(benthos_clean_subs$Year))


time.windowlist=list()
ywindow=5
for(i in 1:(length(yearID)-ywindow)){
time.windowlist[[i]]=  (yearID[i:(i+5)])}

```

# Also crerate time window of 10 year
```{r}
time.windowlist_10=list()
ywindow=10
for(i in 1:(length(yearID)-ywindow)){
time.windowlist_10[[i]]=  (yearID[i:(i+ywindow)])}
```



# how to subset the general data with specific year window
```{r}
benthos_clean_subs %>% 
  filter(Year %in% time.windowlist[[1]])
```

# create time window ID vector (maybe for plotting)
```{r}

tw_ID<-rep(NA, 33)

for(tw in 1:length(time.windowlist)){
  tw_ID[tw]<-

  paste(time.windowlist[[tw]][1], time.windowlist[[tw]][6], sep="_")
}
```

# window 10y ID
```{r}
tw_ID_10<-rep(NA, 28)

for(tw in 1:length(time.windowlist_10)){
  tw_ID_10[tw]<-

  paste(time.windowlist_10[[tw]][1], time.windowlist_10[[tw]][10], sep="_")
}
```



# create a SiteID vector (overall)
```{r}
SiteID_vector<-unique(benthos_clean_subs$SiteID)
```

# site ID vector for sites with >10 observations
```{r}
SiteID_10y<-unique(benthos_clean_subs %>% 
                     filter(SiteID %in% Site_date_counts[Site_date_counts$date_count>10,]$SiteID) %>%
                     pull(SiteID)
)
```




# Prove Synchrony on given time window, filtering and selecting only sites with >4 years in given window;
# Exclude site / year with only one taxa
#include only sites with >4 years data in a given window
# include only sites with >6 taxa in given window
```{r}

codyn::synchrony(
benthos_clean_subs %>% 
                   #distinct() %>% 
                  filter(SiteID %in% Site_date_counts[Site_date_counts$date_count>10,]$SiteID) %>% 
                   filter(Year %in% time.windowlist[[8]]) %>%  # select time window
                   group_by(SiteID, Year) %>% 
                  mutate(taxa_site_year=n_distinct(Taxon.harm)) %>% # identify sites that for a given year only have 1 taxa
                  filter(taxa_site_year>1) %>% # exclude the site - year with only 1 taxa
                  group_by(SiteID) %>% 
                  mutate(nyears= n_distinct(Year)) %>%  # number of year with data for each site
                  filter(nyears>=4) %>% # include sites with >= 4 years data in a given window (of 6y)
                  group_by(SiteID) %>% 
                  mutate(count_taxa=n_distinct(Taxon.harm)) %>% 
                  filter(count_taxa>6) %>% # only include sites with > 6 taxa in a given window
                 group_by(SiteID, Year) %>% 
                  mutate(taxa_site_year=n_distinct(Taxon.harm)) %>% 
                  filter(taxa_site_year>1)

,

time.var = 'Date_julian', species.var = 'Taxon.harm', abundance.var = 'Value', replicate.var = 'SiteID')

```

# *Tests to understand issues with sites with unique species, etc*
```{r}
xx<-
benthos_clean_subs %>% 
                   #distinct() %>% 
                  filter(SiteID %in% Site_date_counts[Site_date_counts$date_count>10,]$SiteID) %>% 
                   filter(Year %in% time.windowlist[[8]]) %>% 
                  select(SiteID, Taxon.harm, Value, Year, Date, Date_julian) %>% 
                  group_by(SiteID, Year) %>% 
                  mutate(taxa_site_year=n_distinct(Taxon.harm)) %>% 
                  filter(taxa_site_year>1) %>% 
                  group_by(SiteID) %>% 
                  mutate(nyears= n_distinct(Year)) %>% 
                  filter(nyears>=4) %>% 
                  group_by(SiteID) %>% 
                  mutate(count_taxa=n_distinct(Taxon.harm)) %>% 
                  filter(count_taxa>6) 
               
              
codyn::synchrony(xx,
time.var = 'Date_julian', species.var = 'Taxon.harm', abundance.var = 'Value', replicate.var = 'SiteID') 

xx %>% 
  group_by(SiteID, Year) %>% 
  summarise(spp_year=n_distinct(Taxon.harm)) %>% arrange(spp_year)


xx %>% 
  filter(SiteID =='SE633738-142203|profundal')


```




# create list holding the moving window synch calculation
```{r}
synch.list<-list()
```

# *Moving window list of comm synchrony*
```{r}
for (tw in 1: length(time.windowlist)) {
  synch.list[[tw]]<-
    
    codyn::synchrony(
benthos_clean_subs %>% 
                   #distinct() %>% # this exclude replicates (canada)
                  #filter(Sampled.habitat != 'Profundal') %>% 
                  filter(SiteID %in% Site_date_counts[Site_date_counts$date_count>10,]$SiteID) %>% # sites with > 10 observations
                   filter(Year %in% time.windowlist[[tw]]) %>% # select given time window
                  group_by(SiteID, Year) %>% 
                  mutate(taxa_site_year=n_distinct(Taxon.harm)) %>% 
                  filter(taxa_site_year>1) %>%                 
                  group_by(SiteID) %>% 
                  mutate(nyears= n_distinct(Year)) %>% 
                  filter(nyears>=4) %>% # at least 4 years in a given window (of 6)
                  group_by(SiteID) %>% 
                  mutate(count_taxa=n_distinct(Taxon.harm)) %>% 
                  filter(count_taxa>6) # sites with at least 6 taxa
                    
,

time.var = 'Date_julian', species.var = 'Taxon.harm', abundance.var = 'Value', replicate.var = 'SiteID')
}
```


```{r}
synch.list[2]
```

# combine elements of list *using reduce (like cbind)*, keeping all siteID and adding NA when a given site has no value in a given timewindow. Not all sites are present in each element of the list. Using reduce we progressively cbind the data keeping all sites, and when a given site is absent (for a given time window) it will produce NAs
```{r}
synch_mw_test<- # call it as a test for now
synch.list %>% 
reduce(full_join, by = "SiteID")

colnames(synch_mw_test)[2:ncol(synch_mw_test)]<-tw_ID
```

# add Country to the synchrony time window
```{r}
synch_mw_test$Country<-
  benthos_clean_subs$Country[match(synch_mw_test$SiteID, benthos_clean_subs$SiteID)]

(synch_mw_test$Country)
```

## Iceland with very large densities
```{r}
benthos_clean_subs %>% 
  filter(SiteID =='Elliðaár')

benthos_clean_subs %>% 
  filter(Country=='Iceland')

benthos_clean_subs %>% 
  filter(ValueType=='Density') %>% 
  ggplot()+aes(x=Country, y = Value)+geom_boxplot()



```


# Transform the synch mw in long format (for plotting)
```{r}
synch_mw_test_long<-
synch_mw_test %>% 
  pivot_longer(cols = -c(SiteID, Country), names_to = 'Year_window', values_to = 'Com.Synchrony')
```

# Plot moving window community synchrony (all country)
```{r}
synch_mw_test_long %>% 
  group_by(SiteID) %>% 
  mutate(na_count = sum(is.na(Com.Synchrony))) %>% #count the NA (for the sites with now synch in given window)
  filter(na_count<20) %>% # exclude sites with > n windows missing
  #mutate(Year_window_n=as.numeric(as.factor(Year_window))) %>% 
  separate(Year_window, into = c("start", "end"), sep = "_", convert = TRUE) %>% 
  mutate(Year_Midpoint = (start + end) / 2) %>% 
  ggplot()+aes(Year_Midpoint, Com.Synchrony, col=Country)+geom_smooth(se=F)

# Iceland is weird
synch_mw_test_long %>% 
  filter(Country!='Iceland') %>% 
  group_by(SiteID) %>% 
  mutate(na_count = sum(is.na(Com.Synchrony))) %>% #count the NA (for the sites with now synch in given window)
  filter(na_count<20) %>% # exclude sites with > n windows missing
  #mutate(Year_window_n=as.numeric(as.factor(Year_window))) %>% 
  separate(Year_window, into = c("start", "end"), sep = "_", convert = TRUE) %>% 
  mutate(Year_Midpoint = (start + end) / 2) %>% 
  ggplot()+aes(Year_Midpoint, Com.Synchrony, col=Country)+geom_smooth(se=F)


```

```{r}
synch_mw_test_long %>% 
  filter(Country!='Iceland') %>% 
  group_by(SiteID) %>% 
  mutate(na_count = sum(is.na(Com.Synchrony))) %>% #count the NA (for the sites with now synch in given window)
  filter(na_count<20) %>% # exclude sites with > n windows missing
  #mutate(Year_window_n=as.numeric(as.factor(Year_window))) %>% 
  separate(Year_window, into = c("start", "end"), sep = "_", convert = TRUE) %>% 
  mutate(Year_Midpoint = (start + end) / 2) %>% 
  ggplot()+aes(Year_Midpoint, Com.Synchrony, col=Country)+geom_smooth(aes(group=SiteID), se=F, size=0.1)+
  geom_smooth(aes(group=Country), se=F)+facet_wrap(~Country)
```

```{r}
table(benthos_family_subs$Lake.river)
```


```{r}
benthos_fam_synchrony 
```


# Plot the number of sites within each moving window
```{r}
synch_mw_test_long %>% 
  drop_na() %>% 
  group_by(Year_window, Country) %>% 
  summarise(n_sites=n_distinct(SiteID)) %>% 
  ggplot()+aes(Year_window, n_sites)+geom_point()+facet_wrap(~Country)+
  theme(axis.text.x = element_text(angle = 90, hjust=-1, size=8))
                                                                                                        
```

# mapping the coords
```{r}
benthos_fam_synchrony %>% 
  filter(Country %in% c('Sweden', 'Norway', 'Finland')) %>% 
  ggplot()+aes(Lon, Lat, col=synchrony)+geom_point()+scale_color_gradient(low='blue', high='red')
```


##########################
## *Moving window Stability*
# create list holding the moving window stability calculation
```{r}
stability.list<-list()
```

```{r}
for (tw in 1: length(time.windowlist)) {
  stability.list[[tw]]<-
    
    codyn::community_stability(
benthos_clean_subs %>% 
                   #distinct() %>% # this exclude replicates (canada)
                  #filter(Sampled.habitat != 'Profundal') %>% 
                  filter(SiteID %in% Site_date_counts[Site_date_counts$date_count>10,]$SiteID) %>% # sites with > 10 observations
                   filter(Year %in% time.windowlist[[tw]]) %>% # select given time window
                  group_by(SiteID, Year) %>% 
                  mutate(taxa_site_year=n_distinct(Taxon.harm)) %>% 
                  filter(taxa_site_year>1) %>%                 
                  group_by(SiteID) %>% 
                  mutate(nyears= n_distinct(Year)) %>% 
                  filter(nyears>=4) %>% # at least 4 years in a given window (of 6)
                  group_by(SiteID) %>% 
                  mutate(count_taxa=n_distinct(Taxon.harm)) %>% 
                  filter(count_taxa>6) # sites with at least 6 taxa
,

time.var = 'Date_julian', abundance.var = 'Value', replicate.var = 'SiteID')
}
```

```{r}
stability.list[[1]]
```

# Combine (reduce) the stability moving window
```{r}
stability_mw_test<- # call it as a test for now
stability.list %>% 
reduce(full_join, by = "SiteID")

colnames(stability_mw_test)[2:ncol(stability_mw_test)]<-tw_ID
```
# add the Country info on the stability moving window
```{r}
stability_mw_test$Country<-
  benthos_clean_subs$Country[match(stability_mw_test$SiteID, benthos_clean_subs$SiteID)]



```


# Transform the stability mw in long format (for plotting)
```{r}
stability_mw_test_long<-
stability_mw_test %>% 
  pivot_longer(cols = -c(SiteID, Country), names_to = 'Year_window', values_to = 'Com.Stability')
```

# Plot of moving window stability per country
```{r}
stability_mw_test_long %>% 
  group_by(SiteID) %>% 
  mutate(na_count = sum(is.na(Com.Stability))) %>% #count the NA (for the sites with now synch in given window)
  filter(na_count<20) %>% # exclude sites with > n windows missing
  #mutate(Year_window_n=as.numeric(as.factor(Year_window))) %>% 
  separate(Year_window, into = c("start", "end"), sep = "_", convert = TRUE) %>% 
  mutate(Year_Midpoint = (start + end) / 2) %>% 
  ggplot()+aes(Year_Midpoint, Com.Stability, col=Country)+geom_smooth(aes(group=SiteID), se=F, size=0.1)+
  geom_smooth(aes(group=Country), se=F)+facet_wrap(~Country)
```

```{r}
stability_mw_test_long %>% 
  group_by(SiteID) %>% 
  mutate(na_count = sum(is.na(Com.Stability))) %>% #count the NA (for the sites with now synch in given window)
  filter(na_count<20) %>% # exclude sites with > n windows missing
  #mutate(Year_window_n=as.numeric(as.factor(Year_window))) %>% 
  separate(Year_window, into = c("start", "end"), sep = "_", convert = TRUE) %>% 
  mutate(Year_Midpoint = (start + end) / 2) %>% 
  ggplot()+aes(Year_Midpoint, Com.Stability, col=Country)+geom_smooth(aes(group=SiteID), se=F, size=0.1)+
  geom_smooth(aes(group=Country), se=F)+facet_wrap(~Country, scale='free_y')
```




# *How many samples per year in each site*
*most sites visited only once per year*
```{r}

Site_samples_year<-
benthos_clean_subs %>% 
  group_by(SiteID, Year) %>% 
  summarise (n_distinct(Date_julian))

#most samples in Fall, some in Spring
benthos_clean_subs %>% 
  group_by(SiteID, Year) %>% 
  summarise (mean_month=mean(Month)) %>% 
  pull(mean_month) %>% 
  hist()

```

# Check distribution of species across sites
*needed to select taxa for spatial synchrony*
```{r}
Taxon_distribution<-
benthos_clean_subs %>% 
  group_by(Taxon.harm) %>% 
  summarise(Distribution=n_distinct(SiteID))
```


###########################################
##### Spatial Synchrony section ########### *perhaps remove profundal sites*

#################################
# *Get grouped table with info for each taxa*
```{r}
benthos_taxa_grp<-
benthos_clean_subs %>%
  group_by(Taxon.harm) %>% 
  nest() %>% 
  ungroup()
  

```

#add the distribution across sites for each taxa
```{r}
benthos_taxa_grp$distribution<-
  Taxon_distribution$Distribution[match(benthos_taxa_grp$Taxon.harm, Taxon_distribution$Taxon.harm)]
```

# have a look at the species 1 in the grouped data
*this df can be used to eventually loop over time windows, for spatial synch moving window*
```{r}
benthos_taxa_grp$data[40]
```


# add a list column, where I create a df with SiteID x Year for each species (if many years, then mean)
*this creates a Site X Year matrix for each species*
```{r}
benthos_taxa_grp<-
benthos_taxa_grp %>% 
  mutate( taxa_wide = map(data, ~ .x %>% # convert to wide format (site x year) Also, here is where we could select time windows
                            select(SiteID, Year, Value) %>% 
                            pivot_wider(names_from = Year,
                                        names_sort = T, # sort the year column
                                        values_from = Value, values_fn = mean))
            )
```

# have a look at the wide format for a taxa
```{r}
benthos_taxa_grp$taxa_wide[[4]] 
 
```

## *Subset the taxa with >20 sites distribution (overall)*
```{r}
benthos_taxa_grp_20sites<- 
  benthos_taxa_grp %>% 
  filter(distribution >20)
```

## actual spatial cor analyses 
# get the *spatial corr (spearman) between each site* (regardless where they are) for all spp in >20 sites overall

```{r}
benthos_taxa_grp_20sites<-
benthos_taxa_grp_20sites %>% 
  mutate(
   spatial_cor = map(taxa_wide, function(df) {
      # 1. Move SiteID to row names so it doesn't get treated as numeric data
      # 2. Transpose so Sites = Columns, Years = Rows
      mat <- df %>%
        column_to_rownames("SiteID") %>%
        t()
      
      # 3. Calculate Spearman Correlation
      # use = "pairwise.complete.obs" handles NAs if sites have different year ranges
      cor(mat, method = "spearman", use = "pairwise.complete.obs")
    })
  )
```



# some species with very sparce matrices
```{r}
benthos_taxa_grp_20sites$taxa_wide[2]
```

# check how many obs in each pairwise
```{r}

# the wide spp data
xx<-
   benthos_taxa_grp_20sites$taxa_wide[[2]]%>%
        column_to_rownames("SiteID") %>%
        t()

# masking presences and NAs
tmp1<-!is.na(xx)
storage.mode(tmp1)<-'numeric'

#matrix of pairwise obs
n_matrix <- t(tmp1) %*% tmp1

# convert to 3-col df, (to add to the pairwise synchrony)
dist.3col(as.dist(n_matrix))



```




# Convert species-specific spatial cor matrix into a 3-column df. Add it as additional column in the group df
```{r}
benthos_taxa_grp_20sites <-
benthos_taxa_grp_20sites %>% 
  mutate(Dist3Cols = map(spatial_cor, function(mat){
    iCAMP::dist.3col(mat) } )
  )
```

# add the Taxon ID in front of the 3-column spatial correlation
```{r}

benthos_taxa_grp_20sites<-
  
benthos_taxa_grp_20sites %>% 
  mutate(Dist3Cols=map2(Dist3Cols, Taxon.harm, ~.x %>% 
                          mutate(TaxaID = .y) %>% 
                          select(TaxaID, everything())
  )
  )
```

# Check that taxa names are now included in
```{r}
benthos_taxa_grp_20sites$Dist3Cols[[2]]
```

# function to *extrct the n observations in each pairwise sites for each spp*
```{r}

n_obs_fun<-function(mat){
  mat=
    mat %>% 
    column_to_rownames("SiteID") %>%
        t()
  mat_01= !is.na(mat)
  storage.mode(mat_01)<-'numeric'
  
  n_matrix=t(mat_01) %*% mat_01
  
  pair_obs=iCAMP::dist.3col(n_matrix)
  
  return(pair_obs)
}


# test the function
n_obs_fun(benthos_taxa_grp_20sites$taxa_wide[[2]])



```




# calculate the number of common (shared) observation between sites, and add this to the grouped df
*also add the name of taxa in the obs pair data*
```{r}
benthos_taxa_grp_20sites<-
  
  benthos_taxa_grp_20sites %>% 
  mutate(obs_pairs = map(taxa_wide, n_obs_fun)) %>% 
  
mutate(obs_pairs=map2(obs_pairs, Taxon.harm, ~.x %>% 
                          mutate(TaxaID = .y) %>% 
                          select(TaxaID, everything())
  )
  )



```

#veryfy the obs.pair data
```{r}
benthos_taxa_grp_20sites$obs_pairs[[1]]
```


```{r}
benthos_taxa_grp_20sites
```



# Create comprehensive 3-col spatial synch across all sites amd species (present in >20 sites)
```{r}
Spatial_synch_all<-
benthos_taxa_grp_20sites %>% 
  select(Dist3Cols) %>% 
  unnest()
```

# add names of County for both sites  on the 3-col spatial synch df
```{r}
Spatial_synch_all$Country1<-
  benthos_clean_subs$Country[match(Spatial_synch_all$name1, benthos_clean_subs$SiteID)]

Spatial_synch_all$Country2<-
  benthos_clean_subs$Country[match(Spatial_synch_all$name2, benthos_clean_subs$SiteID)]


names(Spatial_synch_all)[4]<-'Spatial_synch'

Spatial_synch_all$site_pair<-
  paste(Spatial_synch_all$name1, Spatial_synch_all$name2, sep='_')

#add the combination of site_pair and taxa
Spatial_synch_all$site_pair_TaxaID<-
  paste(Spatial_synch_all$site_pair, Spatial_synch_all$TaxaID, sep="_")

```

# extract the number of observation pairs (site pairs) for each species
#add site pair and site_pair_taxa to the observation pair
```{r}
Spatial_synch_obs.pairs<-
benthos_taxa_grp_20sites %>% 
  select(obs_pairs) %>% 
  unnest()

# add actual pairs of site ID
Spatial_synch_obs.pairs$site.pairs<-
  paste(Spatial_synch_obs.pairs$name1, Spatial_synch_obs.pairs$name2, sep="_")

names(Spatial_synch_obs.pairs)[4]<-'obs.pairs'

Spatial_synch_obs.pairs$site.pairs_TaxaID<-
  paste(Spatial_synch_obs.pairs$site.pairs, Spatial_synch_obs.pairs$TaxaID, sep="_")

```


# add the number of observation pairs for each species

*Now add the obs.pair by matchin site_pair_TaxaID (the complete pairwise info)*
```{r}


Spatial_synch_all$obs.pairs<-
  Spatial_synch_obs.pairs$obs.pairs[match(Spatial_synch_all$site_pair_TaxaID, Spatial_synch_obs.pairs$site.pairs_TaxaID )]
  
  
```


# extract coordinates for *Atlantic sites*
```{r}
SiteID_coords_Atl<-
  benthos_clean_synchrony %>% 
  filter(Country %in% c('Sweden', 'Finland', 'Norway')) %>% 
  select(SiteID, Lon, Lat)
  as.matrix()
  

SiteID_coords_Atl
  
```

#Distance of the Atlantic sites
```{r}
SiteID_geo_dist_Alt<-
  distm(SiteID_coords_Atl[,-1], fun = distGeo) 
  
  #add siteID into the dist matrix
rownames(SiteID_geo_dist_Alt)<-SiteID_coords_Atl$SiteID
colnames(SiteID_geo_dist_Alt)<-SiteID_coords_Atl$SiteID

# convert to Km dist
SiteID_geo_dist_Alt<-
SiteID_geo_dist_Alt/1000


```

#Convert geodist *Atlantic* into 3col df
```{r}
Geodist_Atl_3col<-
dist.3col(SiteID_geo_dist_Alt)

#add site pair information
Geodist_Atl_3col$Site_pair<-
  paste(Geodist_Atl_3col$name1, Geodist_Atl_3col$name2, sep="_")
```


#Add the goedist to the spatial synch df
*this adds geodist only for the Atlantic sites*
```{r}
Spatial_synch_all$GeoDist<-
  Geodist_Atl_3col$dis[match(Spatial_synch_all$site_pair, Geodist_Atl_3col$Site_pair)]
```



# Plot spatial decay of synchrony
#Only Atlantic sites (other have no Geodist)
# Also, seelct only species with >9 observation pairs
```{r}
Spatial_synch_all %>% 
  drop_na() %>% 
  filter(obs.pairs>9) %>% 
  filter(GeoDist <500 & GeoDist>0) %>% 
  ggplot()+aes(sqrt(GeoDist), Spatial_synch)+geom_smooth(col='darkred', size=2)


Spatial_synch_all %>% 
  drop_na() %>% 
  filter(obs.pairs>9) %>% 
  ggplot()+aes(sqrt(GeoDist), Spatial_synch)+geom_smooth(aes(group=TaxaID), size=0.2, se=F, col='grey50')+
  geom_smooth(se=T, col='darkred', size=2)+xlab('Geodist sqrt(km)') + ggtitle('Atlantic sites')


Spatial_synch_all %>% 
  drop_na() %>% 
  filter(obs.pairs>9) %>% 
  filter(GeoDist <200) %>% 
  filter(GeoDist>0) %>% 
  ggplot()+aes(sqrt(GeoDist), Spatial_synch)+geom_smooth(aes(group=TaxaID), size=0.2, se=F, col='grey50')+
  geom_smooth(se=T, col='darkred', size=2)+xlab('Geodist sqrt(km)') + ggtitle('Atlantic sites')


Spatial_synch_all %>% 
  drop_na() %>% 
  filter(TaxaID != 'Oligochaeta') %>% 
  filter(obs.pairs>9) %>% 
  filter(GeoDist <200) %>% 
  filter(GeoDist>0) %>% 
  ggplot()+aes(sqrt(GeoDist), Spatial_synch)+geom_smooth(aes(group=TaxaID), size=0.2, se=F, col='grey50')+
  geom_smooth(se=T, col='darkred', size=2)+xlab('Geodist sqrt(km)') + ggtitle('Atlantic sites')





```


################################
# *Total abundance time series for sites*
```{r}
benthos_clean_tot_abund<-
benthos_clean_subs %>% 
  group_by(SiteID,Year, Lat, Lon, Country, Sampled.habitat, Unit) %>% 
  summarise (CommAbund=sum(Value)) %>% ungroup()
```

# Calculate *total abundance spatial synchrony*
```{r}
Spatial_synch_tot_abund<-
benthos_clean_tot_abund %>% 
    select(SiteID, Year, CommAbund) %>% 
    pivot_wider(names_from = Year,
    names_sort = T, # sort the year column
    values_from = CommAbund) %>% 
   #select(-c(Lat, Lon, Country, Sampled.habitat)) %>% 
  
  column_to_rownames("SiteID") %>%
        t() %>% 
      
      # 3. Calculate Spearman Correlation
      # use = "pairwise.complete.obs" handles NAs if sites have different year ranges
      cor(., method = "spearman", use = "pairwise.complete.obs") %>% 
  iCAMP::dist.3col()


## add site pair info
Spatial_synch_tot_abund$site_pair<-
  paste(Spatial_synch_tot_abund$name1, Spatial_synch_tot_abund$name2, sep="_")
```
            
#get the number of pairwise observations for the tot abundance synchrony
```{r}

Spatial_synch_tot_abund.obs.pairs<-
benthos_clean_tot_abund %>% 
    select(SiteID, Year, CommAbund) %>% 
    pivot_wider(names_from = Year,
    names_sort = T, # sort the year column
    values_from = CommAbund) %>% ungroup() %>% 
   select(-c(Lat, Lon, Country, Sampled.habitat)) %>% 
  n_obs_fun(). # use the previously created function

#add site pair info
Spatial_synch_tot_abund.obs.pairs$site_pairs<-
  paste(Spatial_synch_tot_abund.obs.pairs$name1,Spatial_synch_tot_abund.obs.pairs$name2, sep='_')

names(Spatial_synch_tot_abund.obs.pairs)[3]<-'obs.pairs'

```

# add the n.obs in the tot abund spatial synch df
NOTE: as oppose to the spatial synch of each spp, here the match is based only on site_pair, not site_pair_taxa
```{r}
Spatial_synch_tot_abund$obs.pairs<-
  Spatial_synch_tot_abund.obs.pairs$obs.pairs[match(Spatial_synch_tot_abund$site_pair, Spatial_synch_tot_abund.obs.pairs$site_pairs)]
```


## add geodistance to the total abundance spatial synch
```{r}
Spatial_synch_tot_abund$Geodist<-
  Spatial_synch_all$GeoDist[match(Spatial_synch_tot_abund$site_pair, Spatial_synch_all$site_pair)]
```

```{r}
Spatial_synch_tot_abund %>% 
  drop_na() %>% # remove NAs, which means exlcude non Atlantic sites
  filter(obs.pairs >9) %>% 
  filter(Geodist<500) %>% 
  ggplot()+aes(sqrt(Geodist), dis)+geom_smooth()+ylab('Comm. abundance synchrony')
```

```{r}
Spatial_synch_tot_abund 
```



#### #################################
## *Spatial synchrony for moving window*
####################################

#Pseudocode function

Take data
group data by taxon
add taxa distribution (global)
filter for the given time window [tw] - this is the loop 
mutate to taxa_wide
calculate spatial sync (as cor)
convert matrix to synch 3cols + add site_pair ID
run the obs.pair function + add site_pair ID
Add site pair numbers on the synch 3cols (match using both site_pair and taxaID)
return synch 3 cols


#######################

# create a list to hold the time window of tatal abundance spatial synch
```{r}
abund_spatial_synch_list<-list()
```

# the loop to get time window abundance spat synch
```{r}

for(tw in 1:length(time.windowlist)){

abund_spatial_synch_list[[tw]]=
  
benthos_clean_tot_abund %>% 
  filter(Sampled.habitat != 'Profundal') %>% # remove profundal
    select(SiteID, Year, CommAbund) %>% 
  filter(Year %in% time.windowlist[[tw]]) %>% 
    pivot_wider(names_from = Year,
    names_sort = T, # sort the year column
    values_from = CommAbund) %>%
  
  column_to_rownames("SiteID") %>%
        t() %>% 
      
      # 3. Calculate Spearman Correlation
      # use = "pairwise.complete.obs" handles NAs if sites have different year ranges
      cor(., method = "spearman", use = "pairwise.complete.obs") %>% 
  iCAMP::dist.3col() %>% 
  mutate(year_window=tw_ID[tw]) %>% 
  rename(Synchrony=dis)
}


```

```{r}
abund_spatial_synch_list[[1]]
```


# rbind the *abund spatial synchrony time window*, into a long format
```{r}
abund_spatial_synch_tw<-
rbindlist(abund_spatial_synch_list)

#add the site pair infor
abund_spatial_synch_tw$site_pair<-
  paste(abund_spatial_synch_tw$name1, abund_spatial_synch_tw$name2, sep="_")
```



# get the observation pairs in the time window for abundance synchrony

```{r}
abund.synch_tw_nobspair_list<-list()
```

# the loop to get the number of obs pairs in the abund synch time window
```{r}

for(tw in 1 : length(time.windowlist)) {
  abund.synch_tw_nobspair_list[[tw]]=
benthos_clean_tot_abund %>% filter(Sampled.habitat != 'Profundal') %>% 
    select(SiteID, Year, CommAbund) %>% 
  filter(Year %in% time.windowlist[[tw]]) %>% 
    pivot_wider(names_from = Year,
    names_sort = T, # sort the year column
    values_from = CommAbund) %>% 
  n_obs_fun() %>% 
    rename(n.obs.pairs=dis)
}
  

abund.synch_tw_nobspair_list[[1]]

```

# the long df of number obs pair for the timewindow abund synnchrony
```{r}
abund_spatial_synch_tw_nobspair<-
  rbindlist(abund.synch_tw_nobspair_list)

abund_spatial_synch_tw_nobspair$site_pair<-
  paste(abund_spatial_synch_tw_nobspair$name1, abund_spatial_synch_tw_nobspair$name2, sep="_")
```

#add the observation pair number to the abund synchrony time window df
*Note the max obs.pair == time window of 6 years* That is the number of years contributing to the synch estimates
```{r}
abund_spatial_synch_tw$n.obs.pair<-
  abund_spatial_synch_tw_nobspair$n.obs.pairs[match(abund_spatial_synch_tw$site_pair, abund_spatial_synch_tw_nobspair$site_pair)]
```
# add geodist to the sptial synch tw
```{r}
abund_spatial_synch_tw$Geodist<-
  Spatial_synch_tot_abund$Geodist[match(abund_spatial_synch_tw$site_pair, Spatial_synch_tot_abund$site_pair)]
```

```{r}
abund_spatial_synch_tw<-
abund_spatial_synch_tw %>% 
  mutate(year_window.f=as.factor(year_window)) 
```

```{r}
skipwindow<-levels(abund_spatial_synch_tw$year_window.f)[seq(1, 33, by = 2)]
```

# *Plot of 6years window spatial synchrony patterns in total abundance*
```{r}
abund_spatial_synch_tw %>% 
  drop_na() %>% 
  filter(n.obs.pair>2) %>% 
  filter(year_window.f %in% skipwindow) %>% 
  filter(Geodist<2000) %>% 
  filter(Geodist>0) %>% 
   group_by(site_pair) %>% 
  mutate(nwindows=n_distinct(year_window.f)) %>% 
  filter(nwindows >10) %>% 
  ggplot()+aes(sqrt(Geodist), Synchrony, col=year_window.f) +geom_smooth(method='lm', se=F)+
  scale_color_manual(values= colorRampPalette(c("#5C4033", "#CC7722", "#FFD700"))(17))
```

# *Plot of 6 years mean spatial synchrony over time*
```{r}
abund_spatial_synch_tw %>% 
drop_na() %>% 
  filter(n.obs.pair>3) %>% 
  filter(Geodist>0) %>% 
  filter(Geodist<1000) %>% 
  group_by(site_pair) %>% 
  mutate(nwindows=n_distinct(year_window.f)) %>% 
  filter(nwindows >10) %>% 
  
  group_by(site_pair, year_window.f, Geodist) %>% 
  summarise(mean.spat.sync=mean(Synchrony)) %>% 
   separate(year_window.f, into = c("start", "end"), sep = "_", convert = TRUE) %>% 
  mutate(Year_Midpoint = (start + end) / 2) %>% 
  ggplot()+aes(Year_Midpoint, mean.spat.sync)+geom_smooth() +ylab('Mean spatial synchrony')+ xlab('Year window midpoint')+ggtitle('6 year window')


  
```




##################################################
####### *Time window 10 years* ##### *no profundal*

# create a list to hold the time window of tatal abundance spatial synch
```{r}
abund_spatial_synch_tw10_list<-list()
```

# the loop to get time window abundance spat synch *not profundal sites*
```{r}

for(tw in 1:length(time.windowlist_10)){

abund_spatial_synch_tw10_list[[tw]]=
  
benthos_clean_tot_abund %>% 
  filter(Sampled.habitat != 'Profundal') %>% # remove profundal
    select(SiteID, Year, CommAbund) %>% 
  filter(Year %in% time.windowlist_10[[tw]]) %>% 
    pivot_wider(names_from = Year,
    names_sort = T, # sort the year column
    values_from = CommAbund) %>%
  
  column_to_rownames("SiteID") %>%
        t() %>% 
      
      # 3. Calculate Spearman Correlation
      # use = "pairwise.complete.obs" handles NAs if sites have different year ranges
      cor(., method = "spearman", use = "pairwise.complete.obs") %>% 
  iCAMP::dist.3col() %>% 
  mutate(year_window=tw_ID_10[tw]) %>% 
  rename(Synchrony=dis)
}


```

# there are more site pairs within the 10 years window, but many NAs are present in the overall synchrony df
```{r}
abund_spatial_synch_tw10_list[[1]]


```


# rbind the *abund spatial synchrony time window*, into a long format
```{r}
abund_spatial_synch_tw10<-
rbindlist(abund_spatial_synch_tw10_list)

#add the site pair infor
abund_spatial_synch_tw10$site_pair<-
  paste(abund_spatial_synch_tw10$name1, abund_spatial_synch_tw10$name2, sep="_")
```



# get the observation pairs in the time window 10 for abundance synchrony

```{r}
abund.synch_tw10_nobspair_list<-list()
```

# the loop to get the number of obs pairs in the abund synch time window
```{r}

for(tw in 1 : length(time.windowlist_10)) {
  abund.synch_tw10_nobspair_list[[tw]]=
benthos_clean_tot_abund %>% filter(Sampled.habitat != 'Profundal') %>% 
    select(SiteID, Year, CommAbund) %>% 
  filter(Year %in% time.windowlist_10[[tw]]) %>% 
    pivot_wider(names_from = Year,
    names_sort = T, # sort the year column
    values_from = CommAbund) %>% 
  n_obs_fun() %>% 
    rename(n.obs.pairs=dis)
}
  

abund.synch_tw10_nobspair_list[[1]]

```

# the long df of number obs pair for the timewindow abund synnchrony
```{r}
abund_spatial_synch_tw10_nobspair<-
  rbindlist(abund.synch_tw10_nobspair_list)

abund_spatial_synch_tw10_nobspair$site_pair<-
  paste(abund_spatial_synch_tw10_nobspair$name1, abund_spatial_synch_tw10_nobspair$name2, sep="_")
```

#add the observation pair number to the abund synchrony time window df
*Note the max obs.pair == time window of 10 years* That is the number of years contributing to the synch estimates
```{r}
abund_spatial_synch_tw10$n.obs.pair<-
  abund_spatial_synch_tw10_nobspair$n.obs.pairs[match(abund_spatial_synch_tw10$site_pair, abund_spatial_synch_tw10_nobspair$site_pair)]
```

# add the Geodist (for now only available for Atlantic sites)
```{r}
abund_spatial_synch_tw10$Geodist<-
  Spatial_synch_tot_abund$Geodist[match(abund_spatial_synch_tw10$site_pair, Spatial_synch_tot_abund$site_pair)]
```

```{r}
abund_spatial_synch_tw10<-
abund_spatial_synch_tw10 %>% 
  mutate(year_window.f=as.factor(year_window)) 
```

```{r}
skipwindow10<-levels(abund_spatial_synch_tw10$year_window.f)[seq(1, 28, by = 2)]
```

# *Plot of spatial synchrony patterns 10years window*
```{r}
abund_spatial_synch_tw10 %>% 
  drop_na() %>% 
  filter(n.obs.pair>3) %>% 
  filter(year_window.f %in% skipwindow10) %>% 
  filter(Geodist<1000) %>% 
  filter(Geodist>0) %>% 
  group_by(site_pair) %>% 
  mutate(nwindows=n_distinct(year_window.f)) %>% 
  filter(nwindows >10) %>% 
  
  ggplot()+aes(sqrt(Geodist), Synchrony, col=year_window.f) +geom_smooth(method='lm', se=F)+
  scale_color_manual(values= colorRampPalette(c("#5C4033", "#CC7722", "#FFD700"))(14))+ggtitle('10 year window')


```
```{r}
n_distinct(abund_spatial_synch_tw10$year_window.f)
```

# Plot of spatial synch abund over time for 10years window
```{r}
abund_spatial_synch_tw10 %>% 
drop_na() %>% 
  filter(n.obs.pair>5) %>% 
  filter(Geodist>0) %>% 
  filter(Geodist<1000) %>% 
  group_by(site_pair) %>% 
  mutate(nwindows=n_distinct(year_window.f)) %>% 
  filter(nwindows >10) %>% 
  group_by(site_pair, year_window.f, Geodist) %>% 
  summarise(mean.spat.sync=mean(Synchrony)) %>% 
   separate(year_window.f, into = c("start", "end"), sep = "_", convert = TRUE) %>% 
  mutate(Year_Midpoint = (start + end) / 2) %>%
  ggplot()+aes(Year_Midpoint, mean.spat.sync)+geom_smooth() +ylab('Mean spatial synchrony')+ xlab('Year window midpoint')+ggtitle('10 year window')


```



## Plot spatial synch 10y window, for beginning and end of time series
```{r}
abund_spatial_synch_tw10 %>% 
  drop_na() %>% 
  filter(n.obs.pair>5) %>% 
  filter(year_window.f %in% tw_ID_10[c(1,2,27,28)]) %>% 
  #filter(year_window.f %in% skipwindow10) %>% 
  filter(Geodist<1000) %>% 
  filter(Geodist>0) %>% 
  group_by(site_pair) %>% 
   mutate(nwindows=n_distinct(year_window.f)) %>% 
  filter(nwindows >3) %>% 
  
  ggplot()+aes(sqrt(Geodist), Synchrony, col=year_window.f) +geom_smooth(method='lm', se=F)+
  scale_color_manual(values= c("#5C4033",'brown4', "#CC7722", "#FFD700"))


```

```{r}
abund_spatial_synch_tw10 %>% 
  drop_na() %>% 
  filter(n.obs.pair>5) %>% 
  filter(Geodist<1500) %>% 
  filter(Geodist>0) %>% 
  filter(year_window.f %in% skipwindow10[4]) %>% 
  ggplot()+aes(sqrt(Geodist), Synchrony)+geom_smooth(method='lm')
```





############ LATITUDE effect ##########

```{r}
library(gtools)
```

# create a site metadata with BIN lat values
```{r}
benthos_sites_metad<-
  benthos_clean_subs %>% 
  select(Country, Lake.river, Sampled.habitat, SiteID, Lon, Lat)


# Bin the LAT based on quantiles
benthos_sites_metad$LatBin<-quantcut(benthos_sites_metad$Lat, 3)

benthos_sites_metad$LatBin.   #. Levels: [48.5,57.9] (57.9,62.3] (62.3,69.9]


# Bin the LAT based on cut ranges
benthos_sites_metad<-
benthos_sites_metad %>% 
  mutate(LatCUT= cut_interval(Lat, n=3, labels=c("Low", "Medium", "High")))
```

# Assign the LAT BIN to the total abundance spatial synchrony (NO timewindow)
```{r}
Spatial_synch_tot_abund$LatBinSite1<-
  benthos_sites_metad$LatBin[match(Spatial_synch_tot_abund$name1, benthos_sites_metad$SiteID)]

Spatial_synch_tot_abund$LatBinSite2<-
  benthos_sites_metad$LatBin[match(Spatial_synch_tot_abund$name2, benthos_sites_metad$SiteID)]


Spatial_synch_tot_abund$LatCUTSite1<-
  benthos_sites_metad$LatCUT[match(Spatial_synch_tot_abund$name1, benthos_sites_metad$SiteID)]

Spatial_synch_tot_abund$LatCUTSite2<-
  benthos_sites_metad$LatCUT[match(Spatial_synch_tot_abund$name2, benthos_sites_metad$SiteID)]



```


```{r}
Spatial_synch_tot_abund %>% 
  drop_na() %>% 
  filter(LatBinSite1 == LatBinSite2) %>% 
  filter(obs.pairs>10) %>% 
   filter(Geodist<1000) %>% 
  ggplot()+aes(LatBinSite1, dis)+ geom_boxplot()


Spatial_synch_tot_abund %>% 
  drop_na() %>% 
  filter(LatBinSite1 == LatBinSite2) %>% 
  filter(obs.pairs>9) %>% 
  filter(Geodist<1000) %>% 
  ggplot()+aes(LatBinSite1, Geodist)+ geom_boxplot()


table(Spatial_synch_tot_abund$LatBinSite1)
  
```


```{r}
Spatial_synch_tot_abund %>% 
  drop_na() %>% 
  filter(LatCUTSite1 == LatCUTSite2) %>% 
  filter(obs.pairs>9) %>% 
   filter(Geodist<1000) %>% 
  ggplot()+aes(LatCUTSite1, dis)+ geom_boxplot()


Spatial_synch_tot_abund %>% 
  drop_na() %>% 
  filter(LatCUTSite1 == LatCUTSite2) %>% 
  filter(obs.pairs>9) %>% 
   filter(Geodist<500) %>% 
  ggplot()+aes(LatCUTSite1, Geodist)+ geom_boxplot()

```




# assign the LAT bin to the abundance synchrony tw10
```{r}
abund_spatial_synch_tw10$LatBinSite1<-
  benthos_sites_metad$LatBin[match(abund_spatial_synch_tw10$name1, benthos_sites_metad$SiteID)]

abund_spatial_synch_tw10$LatBinSite2<-
  benthos_sites_metad$LatBin[match(abund_spatial_synch_tw10$name2, benthos_sites_metad$SiteID)]

```




